The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-06-01 paper.tex
integrate problems and research gaps, still missing some source integration

\section{Background}
There are various nomenclatures for the description of anatomical landmarks.
These allow a uniform classification of anatomical features and structures:

@@

% TODO: add rdfbones

>>>>

\section{Related Work}
There are various nomenclatures for the description of anatomical landmarks.
These allow a uniform classification of anatomical features and structures:

@@

\subsection{Problems and Research Gaps}

There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature - general or subject-specific - can address adequately so far.

The primary issue lies in the absence of standardisation. While both the TA and FMA have been proposed as a means of standardising terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardisation in actual practice~\citep{frequencyta,doestamatter,athighlights}. 
Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.  
Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas. As a consequence there are at times various translations for the same anatomical terms. Yet, knowledge of Latin is generally declining. With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes.
Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes. For instance <<{{Quelle}}> uses the German terms of the bones in the explanation of sutures. 
Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example,  there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure. However, this kind of naming variable, non-intuitive and therefore inconsistent. 
An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in {{Quelle}} is described as saddle-shaped although it is not called ‚sella‘, the Latin name for saddle. It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.  
The Cuspis dentis, the tooth cusp, can by shape also be described as a protuberance. The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination. 
As can be seen, the incentive for synonyms is high. However, the current resources fail to provide comprehensive lists of synonyms in use.  This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the „preferred name„ External acoustic aperture and „non-English equivalent“  Porus acusticus externus (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym. However, it is not the same, rather the Porus is - as its English name indicates - the opening to the meatus.   
Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong. When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures. For instance, the position of the Corpus ossis pubis varies across different sources.

In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.
Another main issue  is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field. For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g. on the Ossa carpi or Ossa metatarsi) are defined in the TA or FMA, resulting in further gaps. 
A similar situation exists with respect to osteometric measurements. Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible.  However, since there are measurement distances on all bones, the respective start and end points must also be defined. Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants.
Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures, which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy . A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other ‘articular eminence’. Often exchange between these specific fields and general anatomy hardly takes place. It stands to reason that interdisciplinary coordination between the individual disciplines and general anatomy hardly takes place.

Furthermore, visual representation of anatomical structures is frequently missing or insufficient in anatomical resources. The labelling is often selective, varies by source and is mostly limited to arrows as a means for local designation. However, since they are  multi-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately. The latter also requires a clear textual definition. 
 However, there are currently no equivalent standardised definitions for skeletal anatomy, meanwhile anatomical resources only provide sparse information on the topic.

For these reasons, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances. Additionally, these should be presented in a clear way within an ontology. The ontology should be easy to understand by the user and include relevant knowledge for the application of the terminology, such as clear descriptions of individual Latin declensions.
